ODGI: understanding pangenome graphs

### 2022-02-21 manuscript/main.tex
Minor comments/fixes

\section{Introduction}
A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{cpang2018,Eizenga_2020}}.
In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented\REVIEWED{~\citep{Ballouz2019}}.
\REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
In node-labeled sequence graphs, nodes indicate \REVIEWED{[1,2] genomic sequences [1,2]}, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of the genomic sequences [1,2]}.
On this model, variation graphs add the concept of paths representing linear \REVIEWED{[1,2] genomic [1,2]} sequences as traversals through the nodes of the graph~\citep{Garrison:2018}. \REVIEWED{[1,2] For example, a [1,2]} path can be a genome, haplotype, contig, or read. [2,5]}

@@

Although \REVIEWED{pangenome graphs} are data structures of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset \REVIEWED{capable of operating on gigabase-scale pangenome graphs constructed from whole-genome assemblies}.
Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not handle efficiently pangenome graphs presenting complex motifs that result from repetitive sequences like centromeres.}
Here we refocus it with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
ODGI offers a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs,
and providing a variety of operations, such as visualization, sorting, and liftover projections, all critical to understand and exploit pangenome graphs.

@@

Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties and shape of the graph.
Graph node order can affect downstream analyses on pangenome graphs.
With \textit{odgi viz} we can color the paths by path position (Fig.~\ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they end.
This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
Pangenome graphs represent both strands \REVIEWED{[1,2] of the genomic sequences [1,2]}.
\textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
With \textit{odgi viz}, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth,} highlighting the different copy number statuses in the \REVIEWED{[1,2] genomic sequences [1,2]} represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).

@@

\REVIEWED{When transforming a GFAv1 with VG, we went for the XG file format, because all downstream tools involved in the evaluation require this static graph representation model in VG.}

\REVIEWED{Our general findings are that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, Tab.~\ref{tab:position}) across all operations. With raising number of haplotypes in the graph, ODGI scales much butter than VG. For example, extracting a subgraph (Fig.~\ref{fig:eval-extract}), ODGI is up to 40 times faster and requires 8 times less memory than VG. Surprisingly, \textit{odgi build} occupies less memory with more threads. We don't have an explanation for that.
}
\REVIEWED{Both visualization tools can only make use of a single thread. For a 1 haplotype graph \textit{vg viz} produces a 816MB SVG which can't be opened by any program. For example, GIMP2 fails with "cannot load more than 1000000 XML elements". For larger graphs, \textit{vg viz} runs through and produces SVGs with only the XML header. This makes it unusable for larger graphs.}

@@

\REVIEWED{The performance analysis shows that ODGI clearly outperforms VG when handling large, complex pangenome graphs. This is one major achievement, because VG's XG data structure is actually static, compared to ODGI's dynamic graph implementation. Across the evaluation of key graph operations, ODGI's memory peak was 10GB. This makes it perfectly suited to be run interactively on a recent laptop. VG has limitations here. We expect that ODGI will be able to handle the next phase of the HPRC, constructing a pangenome graph from 300 individuals, without any problems.}

>>>>

\section{Introduction}
A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{cpang2018,Eizenga_2020}}.
In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented\REVIEWED{~\citep{Ballouz2019}}.
\REVIEWED{[2,5] A class of methods to represent pangenomes involves sequence graphs~\citep{Hein1989, Paten:2017} where homologous regions between genomes are compressed into a single representative \FIXME{representation?} of all alleles present in the pangenome.
In node-labeled \FIXME{?} sequence graphs, nodes indicate \REVIEWED{[1,2] genomic sequences [1,2]} with edges connecting those nodes.
A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of DNA [1,2]}.
On this model, variation graphs add the concept of paths representing linear \REVIEWED{[1,2] DNA [1,2]} sequences as traversals through the nodes of the graph~\citep{Garrison:2018}. \REVIEWED{[1,2] For example, a [1,2]} path can be a genome, haplotype, contig, or read. [2,5]}

@@

Although \REVIEWED{pangenome graphs} are data structures of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset \REVIEWED{capable of operating on gigabase-scale pangenome graphs constructed from whole-genome assemblies}.
Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not efficiently handle pangenome graphs presenting complex motifs that result from repetitive sequences (see Fig.~\ref{fig:1}).}
Here we refocus it \FIXME{the effort?} with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
ODGI offers a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs,
and providing a variety of operations, such as visualization, sorting, and liftover projections, all critical to understand and exploit pangenome graphs.

@@

Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties and shape of the graph.
Graph node order can affect downstream analyses on pangenome graphs.
With \textit{odgi viz} we can color the paths by path position (Fig.~\ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they end.
This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
Pangenome graphs represent both strands \REVIEWED{[1,2] of genomic sequences \FIXME{DNA?}[1,2]}.
\textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
With \textit{odgi viz}, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth,} highlighting the different copy number statuses in the \REVIEWED{[1,2] genomic sequences [1,2]} represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).

@@

\REVIEWED{When transforming a GFAv1 with VG, we used the XG file format because all downstream tools involved in the evaluation required this static graph representation model of VG.} \FIXME{Funny grammar, can be shorter}.

\REVIEWED{Our general findings are that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, Tab.~\ref{tab:position}) across all operations. With raising number of haplotypes in the graph, ODGI scales much butter than VG. For example, extracting a subgraph (Fig.~\ref{fig:eval-extract}), ODGI is up to 40 times faster and requires 8 times less memory than VG. Surprisingly, \textit{odgi build} occupies less memory with more threads. We don't have an explanation for that.
} \FIXME{I would leave that statement out}
\REVIEWED{Both visualization tools can only make use of a single thread. For a 1 haplotype graph \textit{vg viz} produces a 816MB SVG which can't be opened by any program. For example, GIMP2 fails with "cannot load more than 1000000 XML elements". For larger graphs, \textit{vg viz} runs through and produces SVGs with only the XML header. This makes it unusable for larger graphs.}

@@

\REVIEWED{The performance analysis shows that ODGI clearly outperforms VG when handling large, complex pangenome graphs. This is a major achievement, because VG's XG data structure is static, compared to ODGI's dynamic graph implementation. Across the evaluation of key graph operations, ODGI's memory peak was 10GB. This makes it perfectly suited to be run interactively on a recent laptop. VG has limitations here. We expect that ODGI will be able to handle the next phase of the HPRC, constructing a pangenome graph from 300 individuals, without any problems.}
